# Tuesday, August 23, 2022 {-}

* Is it getting hotter here in Tallahassee?
* Are Atlantic hurricanes getting stronger?

Data science (formerly known as 'statistics') is an exciting discipline that allows you to turn data into understanding, insight, and knowledge.

## Today {-}

- Understand what this course is about, how it is structured, and my expectations for you
- Start working with RStudio and R.

## What is this course? {-}

This is designed as first course in data science for geographers.

**Q - What statistics background does this course assume?**  
A - None.

**Q - Is this an intro stat course?**  
A - Statistics and data science are closely related with much overlap. Hence, this course is a great way to get started with statistics. But this course is **not** your typical high school/college statistics course.

**Q - Will we be doing computing?**   
A - Yes.

**Q - Is this an introduction to computer science course?**  
A - No, but many themes are shared.

**Q - What computing language will we learn?**  
A - R.

**Q - Why not language some other language?**  
A - We can discuss that over coffee.

## Where are the materials for this course? {-}

[Github](https://github.com/jelsner/QG-2022)

[RStudio Cloud](https://rstudio.cloud)

## Join RStudio Cloud {-}

* Go to [RStudio Cloud](https://rstudio.cloud), and log in.
* Click on this link https://rstudio.cloud/spaces/75344/join?access_code=drxxMUzCFzMd2hUzGi5EZZaR9CgXY2jJVC6mz54L to join the Quantitative Geography Using R Space on RStudio Cloud.

## Examples {-}

Some of my recent research:

- [More hots](https://eartharxiv.org/q4y8z/)
- [Stronger tornadoes](https://eartharxiv.org/wpkt9/)

Other research:

- [A year as told by fitbit](http://livefreeordichotomize.com/2017/12/27/a-year-as-told-by-fitbit/) by Nick Strayer
- [R-Ladies global tour](http://www.masalmon.eu/2017/10/06/globalrladiestour/) by Maelle Salmon

## Course Syllabus {-}

- Log on to RStudio Cloud and click on this course's Space (Quantitative Geography Using R).
- Click on the project `00_Syllabus` and launch it.
- Open the `00-Syllabus.Rmd` file (lower-right panel), and then click on the "Knit" button.
- Review

## Tallahassee Daily Temperatures {-}

- Packages > Install
- In the Packages window, type tidyverse, lubridate, here, ggplot2 then select Install

Get the data into your environment.
```{r, get the data}
TLH.df <- readr::read_csv(file = here::here('data', 'TLH_SOD1892.csv'),
                          show_col_types = FALSE) |>
      dplyr::filter(STATION == 'USW00093805') |>
      dplyr::mutate(Date = as.Date(DATE)) |>
      dplyr::mutate(Year = lubridate::year(Date), 
             Month = lubridate::month(Date), 
             Day = lubridate::day(Date),
             doy = lubridate::yday(Date)) |>
      dplyr::select(Date, Year, Month, Day, doy, TMAX, TMIN, PRCP)
```

Package name :: function name

Or, load the packages by name with the `library()` function in the file above where they are first used.

Create a plot of the frequency of high temperatures by temperature values.
```{r, create a plot}
library(dplyr)
library(ggplot2)

TLH.df |>
  group_by(TMAX) |>
  summarize(nH = n()) |>
ggplot(mapping = aes(x = TMAX, y = nH)) +
  geom_col(col = 'white', fill = "gray70") +
  labs(title = "Frequency of Daily High Temperatures",
       subtitle = "Tallahassee, FL, USA (1940-2018)",
       x = "Daily High Temperature (°F)",
       y = "Number of Days") +
 scale_x_continuous(breaks = seq(from = 20, to = 110, by = 10)) +
 theme_minimal()
```
