# Tuesday, August 23, 2022 {-}

- Is it getting hotter here in Tallahassee?
- Are Atlantic hurricanes getting stronger?

Data science is a discipline that teaches you how to turn data into understanding, insight, and knowledge. This is a course in data science within the context of geography.

Today

- What this course is about, how it is structured, and what I expect from you
- Getting started with R and RStudio

## What is this course? {-}

This is a course in data science for geographers.

**Q - What statistics background does this course assume?**  
A - None.

**Q - Is this an introductory statistics course?**  
A - Statistics and data science are closely related with much overlap. Hence, this course is a good way to get started with statistics. But this course is **not** a typical high school/college statistics course.

**Q - Will you be doing computing?**   
A - Yes.

**Q - Is this an introduction to computer science course?**  
A - No, but many themes are shared.

**Q - What computing language will you learn?**  
A - R.

**Q - Why not language some other language?**  
A - We can discuss that over coffee in the new Student Union.

## Where are the materials for this course? {-}

[Github](https://github.com/jelsner/QG-2022)

Some of my recent research:

- [More hots](https://eartharxiv.org/q4y8z/)
- [Stronger tornadoes](https://eartharxiv.org/wpkt9/)

Other research:

- [A year as told by fitbit](http://livefreeordichotomize.com/2017/12/27/a-year-as-told-by-fitbit/) by Nick Strayer
- [R-Ladies global tour](http://www.masalmon.eu/2017/10/06/globalrladiestour/) by Maelle Salmon

## Get the syllabus {-}

- Navigate to https://github.com/jelsner/QG-2022
- Click on the file labeled `00-Syllabus.Rmd`
- Click on the button `[Raw]`
- You can `Save as` to download a copy to your computer

## Getting setup {-}

First get R

* Go to http://www.r-project.org.
* Select the CRAN (Comprehensive R Archive Network). Scroll to a mirror site. 
* Choose the appropriate file for your hardware.
* Follow the instructions to install R.

Then get RStudio

* Go to on http://rstudio.org
* Download RStudio Desktop
* Install and open RStudio

Then get the course materials from GitHub

* Navigate to https://github.com/jelsner/QG-2022
* Click the Green Code button
* Then Download ZIP
* Unzip the file on your computer to create the folder `QG-2022-main`

Then open the first two Rmd files

* Open the folder to see all the lesson files (and more)
* Click on the files `00-Syllabus.Rmd` and `01-Lesson.Rmd`

This should bring up the files in RStudio.

Finally, learn how to connect to GitHub (optional). See Jennifer Bryan's https://happygitwithr.com/install-git.html

## Getting to know RStudio's integrated development environment (IDE) {-}

* Written in HTML
* Top menus
  + File > New File > R Markdown
  + Tools > Global Options > Appearance
  
* Upper left panel is the markdown file. This is where you put your text and code. 
  + Run code chunks from this panel 
  + Output from the operations can be placed in this panel or in the Console (see the gear icon above)
  + All the text, code, and output can be rendered to an HTML file or a PDF or Word document (see the Knit button above)
  
* Upper right panel shows what is in your current environment and the history of the commands you issued.
  + This is also where you can connect to github
  
* Lower left panel is the Console
  + I think of this as a sandbox where you try out small bits of code. If it works and is relevant move it to the markdown file.
  + This is also where output from running code will be placed.
  + Not a place for plain text
  
* Lower right panel shows your project files, the plots that get made, and all the packages associated with the project.
  + The File tab shows the files in the project. The most important one is the .Rmd.
  + The Plot tab currently shows a blank sheet
  + The Packages tab shows all the packages that have been downloaded from CRAN and are associated with this project.

Example: Tallahassee daily temperatures

- Packages > Install
- In the Packages window, type tidyverse, lubridate, here, ggplot2 then select Install

Get the data into your RStudio environment.

```{r}
TLH.df <- readr::read_csv(file = here::here('data', 'TLH_SOD1892.csv'),
                          show_col_types = FALSE) |>
      dplyr::filter(STATION == 'USW00093805') |>
      dplyr::mutate(Date = as.Date(DATE)) |>
      dplyr::mutate(Year = lubridate::year(Date), 
             Month = lubridate::month(Date), 
             Day = lubridate::day(Date),
             doy = lubridate::yday(Date)) |>
      dplyr::select(Date, Year, Month, Day, doy, TMAX, TMIN, PRCP)
```


Use package::function  (`::` is called a library specifier). Or, load the package functions into your current RStudio environment with the `library()` function.

Click on the Environment tab to see the data object `TLH.df`.

Use the data object and create a plot showing the frequency of daily high temperatures in Tallahassee.

```{r}
library(ggplot2)

TLH.df |>
  dplyr::group_by(TMAX) |>
  dplyr::summarize(nH = dplyr::n()) |>
ggplot(mapping = aes(x = TMAX, y = nH)) +
  geom_col(col = 'white', fill = "gray70") +
  labs(title = "Frequency of Daily High Temperatures",
       subtitle = "Tallahassee, FL, USA (1940-2018)",
       x = "Daily High Temperature (°F)",
       y = "Number of Days") +
 scale_x_continuous(breaks = seq(from = 20, to = 110, by = 10)) +
 theme_minimal()
```
