# Tuesday, September 13, 2022 {.unnumbered}

Note: Thursday I will be on travel. I will have Assignment #2 available on Canvas by Thursday morning.

Today

-   Data visualization

## Data visualization {.unnumbered}

Data visualization is one of the cornerstones of data science. It allows insights into your data that are not accessible by looking at a spreadsheet or data frame of values.

The {ggplot2} package provides functions to make plots efficiently. The functions are an application of the grammar of graphics theory (Leland Wilkinson) of data visualization.

At a basic level, graphs/plots/charts (all interchangeable terms) provide a way to explore the patterns in data; the presence of extreme values, distributions of individual variables, and relationships between groups of variables.

A graph should emphasize the findings and insights you want your audience to understand. This requires a balance.

On one hand, you want to highlight as many interesting findings as possible. On the other hand, you don't want to include too much information that it overwhelms the audience.

The grammar of graphics specifies how a plot translates data into attributes and geometric objects. - Attributes are things like location on along an axes, color, shape, and size. - Geometric objects are things like points, lines, bars, and polygons.

The type of plot depends on the geometric object, which is specified as a function.

Function names for geometric objects begin with `geom_`. For example, to create a scatter plot of points the `geom_point()` function is used.

Make the functions from the {ggplot2} package available in your current session.

```{r}
library(ggplot2)
```

Bar chart: A simple graph is the bar chart showing the number of cases within each group. Consider again the annual hurricane counts.

Import the data from the file on my website and list the first six rows.

```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/US.txt"
LH.df <- readr::read_table(loc)
dplyr::glimpse(LH.df)
```

Recall that each case is a year and that the function `table()` returns the number of years for each landfall count.

```{r}
table(LH.df$All)
```

The number of cases for each count is tallied and displayed below the count. There were 36 cases (years) with no hurricanes.

The function `geom_bar()` creates a bar chart of this frequency table.

```{r}
ggplot(data = LH.df) + 
  geom_bar(mapping = aes(x = All))
```

You begin a plot with the function `ggplot()` that creates a coordinate system that you add layers to. The first argument of `ggplot()` is the data frame to use in the graph. So `ggplot(data = LH.df)` creates an empty graph.

You complete the graph by adding one or more layers. The function `geom_bar()` adds a layer of bars to our plot, which creates a bar chart.

Each `geom_` function takes a mapping argument. This defines how variables in our data frame are mapped to visual properties. The mapping argument is always paired with `aes()` function, and the `x` argument of `aes()` specifies which variables to map to the x axes, in this case `All`. `ggplot()` looks for the mapped variable in the data frame defined by the data argument, in this case, `LH.df`.

The function `geom_bar()` tables the counts and then maps the number of cases to bars with the bar height proportional to the number of cases. Here the number of cases is the number of years with that many hurricanes.

The functions are applied in order (`ggplot()` comes before `geom_bar()`) and are linked with the addition `+` symbol. In this way you can think of the functions as layers in a GIS.

The bar chart contains the same information as displayed by the function `table()`. The y-axis label is 'count' and x-axis label is the column name.

Repeat this time using Florida hurricane counts. The annual number of Florida hurricanes by year is given in column `FL` in the data frame `LH.df`.

```{r}
LH.df$FL
```

The `geom_bar()` function tables these numbers and plots the frequency as a bar.

```{r}
ggplot(data = LH.df) + 
  geom_bar(mapping = aes(x = FL)) +
  xlab("Number of Florida Hurricanes (1851-2016)") +
  ylab("Number of Years")
```

Here axes labels are placed on the plot with the functions `ylab()` and `xlab()`. With this type of 'layering' it's easy to go from data on the web to a publishable plot.

Example: Pie preference. Thirty graduate students were surveyed about their favorite pie. Categories were (1) chocolate cream, (2) coconut custard, (3) georgia peach, and (4) strawberry.

To make a bar chart showing the number of students who chose each category first create a character vector and then change the vector to a data frame.

```{r}
pie <- c(rep('chocolate cream', times = 4), 
         rep('coconut custard', times =  12), 
         rep('georgia peach', times =  5), 
         rep('strawberry', times =  9))
pie.df <- pie |>
  as.data.frame()
```

Use the function `str()` to see the data type in the data frame.

```{r}
str(pie.df)
```

There is a single variable in the data frame with the name `pie`. The variable is a character string.

Create a table.

```{r}
table(pie.df$pie)
```

Create a bar chart and specify the axis labels.

```{r}
ggplot(data = pie.df) + 
  geom_bar(mapping = aes(x = pie)) +
  xlab("Pie Preference") + 
  ylab("Number of Students")
```

This is a start. Improvements should be made.

First, the bar order is alphabetical from left to right. This is the default ordering for character vectors. It is easier to make comparisons if frequency determines the order.

To change the order on the bar chart (1) convert the variable from character to factor with the `factor()` function then (2) specify the order of the factor levels based on frequency with the `forcats::fct_infreq()` function.

```{r}
pie <- pie |>
  factor() |>
  forcats::fct_infreq()
```

Make the table.

```{r}
table(pie)
```

Now remake the bar chart.

```{r}
pie.df <- pie |>
  as.data.frame()
```

```{r}
ggplot(data = pie.df) + 
  geom_bar(mapping = aes(x = pie)) +
  xlab("Pie Preference") + 
  ylab("Number of Students")
```

Second, the vertical axis tic labels are fractions. Since the bar heights are counts (integers) the tic labels also should be integers.

To override this default you add a new y-axis layer. The layer is the function `scale_y_continuous()` where you indicate the lower and upper limits of the axis with the concatenate (`limits = c()`) function. Now remake the bar chart.

```{r}
ggplot(data = pie.df) + 
  geom_bar(mapping = aes(x = pie)) +
  xlab("Pie Preference") + 
  ylab("Number of Students") +
  scale_y_continuous(limits = c(0, 15))
```

Now the chart is publishable. Options exist for changing the look of the plot for digital media include, colors, orientation, background, etc.

For example to change the bar color use the `fill =` argument in the function `geom_bar()`. To change the orientation of the bars use the layer function `coord_flip`, and to change the background use the layer function `theme_minimal()`. You make changes to the look of the plot with additional layers.

```{r}
ggplot(data = pie.df) + 
  geom_bar(mapping = aes(x = pie), fill = "blue") +
  xlab("Pie Preference") + 
  ylab("Number of Students") +
  scale_y_continuous(limits = c(0, 15)) +
  coord_flip() +
  theme_minimal()
```

Note that the `fill =` is used on the variable named in the `aes()` function but it is specified outside the `aes()` function.

Available colors include

```{r eval=FALSE}
colors()
```

Example: Consider storm intensity of tropical cyclones during 2017. First create two vectors one numeric containing the minimum pressures (millibars) and the other character containing the storm names.

```{r}
minP <- c(990, 1007, 992, 1007, 1005, 
          981, 967, 938, 914, 938, 
          972, 971)
name <- c("Arlene", "Bret", "Cindy", "Don", "Emily", 
          "Franklin", "Gert", "Harvey", "Irma", "Jose", 
          "Katia", "Lee")
```

The function `reorder()` takes a character vector as the first argument and returns an ordered factor with the order dictated by the numeric values in the second argument.

```{r}
name <- name |>
  reorder(minP)
```

The vector `name` is in alphabetically order but the factor levels indicate the order of storms from lowest pressure (Irma) to highest pressure (Don).

Using the `mutate()` function you can add a column to a data frame where the column is an ordered factor.

Note that it is the difference in pressure (`deltaP` for short) between the air outside the tropical cyclone and the air in the center that causes the winds. Cyclones with a large pressure difference are stronger in terms of their wind speed.

Typically the air outside is about 1014 mb so you compute `deltaP` and then reorder the tropical cyclone names using this computed variable.

```{r}
df <- data.frame(name, minP) |>
  dplyr::mutate(deltaP = 1014 - minP,
                nameOrderedFactor = reorder(name, deltaP))
```

Finally you plot the bar chart. Since there is no tabulation of the values you use `geom_col()` instead of `geom_bar()`.

```{r}
ggplot(data = df) + 
  geom_col(mapping = aes(x = nameOrderedFactor, y = deltaP)) +
  ylab("Pressure Difference [mb]") +
  xlab("Atlantic Tropical Cyclones of 2017") +
  coord_flip()
```

Note: `geom_bar()` plots a bar chart AFTER tabulating a column. `geom_col()` plots a bar chart on a pre-tabulated column.

Let's return to the weather data from Tallahassee.

```{r}
df <- readr::read_csv(file = here::here("data", "TLH_SOD1892.csv")) |>
  dplyr::filter(STATION == "USW00093805") |>
  dplyr::mutate(Year = lubridate::year(DATE),
         Month = lubridate::month(DATE)) |>
  dplyr::filter(Year >= 1980 & Month == 9) |>
  dplyr::group_by(Year) |>
  dplyr::summarize(TotalPrecip = sum(PRCP)) |>
  dplyr::mutate(Year = reorder(as.factor(Year), TotalPrecip))

ggplot(data = df) + 
  geom_col(mapping = aes(x = Year, y = TotalPrecip)) +
  ylab("September Rainfall [in]") +
  coord_flip()
```

The histogram is similar to the bar chart except it uses bars to indicate frequency (or proportion) over an interval of continuous values. For instance, with continuous values the function `table()` is not useful.

```{r}
x <- rnorm(n = 10)
table(x)
```

So neither is a bar plot.

A histogram is made as follows: First a collection of disjoint intervals, called bins, covering the range of data points is chosen. "Disjoint" means no overlap, so the intervals look like (a,b] or [a,b). The interval (a,b] means the interval contains all the values from a to b including b but not a, whereas the interval [a,b) means the interval contains all the values from a to b including a but not b.

Second, the number of data values in each of these intervals is counted. Third, a bar is drawn above the interval so that the area of the bar is proportional to the frequency. If the intervals defining the bins have the same width, then the height of the bar is proportional to the frequency (the number of values inside the interval).

Let's return to the Florida precipitation data.

```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt"
FLp.df <- readr::read_table(loc)
```

Recall that the columns in the data frame `FLp.df` are months (variables) and rows are years. `Year` is an integer (`int`) vector and the months are numeric (`num`) vectors. Create a histogram of May precipitation.

```{r}
ggplot(data = FLp.df) + 
  geom_histogram(mapping = aes(x = May), col = "white")  +
  xlab("May Precipitation in Florida (in)") 
```

By default the function `geom_histogram()` picks 30 bins. Since there are only 118 May values many of the bins have fewer than 5 values.

When making a histogram you need to vary the number of bins before deciding on a final plot. This can be done with the `bins =` or `binwidth =` argument. For example, the look of the histogram is improved by halving the default number of bins.

```{r}
ggplot(data = FLp.df) + 
  geom_histogram(mapping = aes(x = May), 
                 col = "white", 
                 bins = 15)  +
  xlab("May Precipitation in Florida (in)") 
```

It looks even better by decreasing the number of bins to 11.

```{r}
ggplot(data = FLp.df) + 
  geom_histogram(mapping = aes(x = May), 
                 col = "white", 
                 bins = 11, 
                 fill = "green3")  +
  xlab("May Precipitation in Florida (in)") +
  ylab("Number of Years")
```

Here the `fill =` argument is used to change color and a `ylab()` layer is added to make the y-axis label more concise.

The `geom_rug()` layer adds the location of the data values as tic marks just above the horizontal axis. And the `color = "white"` is the color of the bin boundaries.

```{r}
ggplot(data = FLp.df) + 
  geom_histogram(mapping = aes(x = May), 
                 col = "white", 
                 bins = 11, 
                 fill = "green3")  +
  xlab("May Precipitation in Florida (in)") +
  ylab("Number of Years") +
  geom_rug(mapping = aes(x = May))

ggplot(data = FLp.df, mapping = aes(x = May)) + 
  geom_histogram(col = "black", 
                 bins = 11, 
                 fill = "pink")  +
  xlab("May Precipitation in Florida (in)") +
  ylab("Number of Years") +
  geom_rug()
```

Density plot

A density plot is a smoothed histogram with units of probability on the vertical axis. It's motivated by the fact that for a continuous variable, the probability that the variable takes on any particular value is 0. Instead you need a range of values over which a probability is defined.

The probability density answers the question, what is the chance that a value falls in a small interval. This chance varies depending on where the value is located within the distribution of all values (e.g., near the middle of the distribution the chance is highest).

```{r}
ggplot(data = FLp.df) +
  geom_density(mapping = aes(x = May)) +
  xlab("May Precipitation in Florida (in)") 
```

The vertical axis is the average chance that rainfall will take on a value along the horizontal axis within a given small interval. The size of the interval is determined by the bandwidth (`bw =`).

The values along the vertical axis depends on the data units. It can be tricky to interpret. Instead `geom_freqpoly()` produces a density-like graph where the units on the y-axis are counts as with the histogram.

```{r}
ggplot(data = FLp.df, aes(x = May)) + 
  geom_freqpoly(color = "green3", binwidth = 1) +
  xlab("May Precipitation in Florida (in)") +
  geom_rug()
```

Box and whisker plot

The box plot graphs the summary statistics. These statistics include the minimum value, the maximum value, the 1st & 3rd quartile values, and the median value. The easiest way to create a box plot is to use the function `boxplot()`.

```{r}
boxplot(FLp.df$May)
```

The function `boxplot()` is from the base {graphics} package. It is not a {ggplot2} function. Others from this package include `hist()` for histograms and `plot()` for scatter plots.

The base graphics lets you manipulate details of a graph. For example:

```{r}
boxplot(FLp.df$May, 
        ylab = "May Precipitation in FL (in)")
f <- fivenum(FLp.df$May)
text(rep(1.3, 5), f, labels = c("Minimum", "1st Quartile", 
                                "Median", "3rd Quartile",
                                "Maximum"))
text(1.3, 7.792, labels = "Last Value Within\n 1.5xIQR Above 3rd Q")
```

The box plot illustrates the five numbers graphically. The median is the line through the box. The bottom and top of the box are the 1st and 3rd quartile values. Whiskers extend vertically from the box downward toward the minimum and upward toward the maximum.

If values extend beyond 1.5 times the interquartile range (either above or below the corresponding quartile) the whisker is truncated at the last value within the range and points are used to indicate outliers.

To make a box plot using the function `ggplot()` you need a dummy variable for the `x` argument in the function `aes()`. This is done with `x = ""`.

```{r}
ggplot(FLp.df) + 
  geom_boxplot(mapping = aes(x = "", y = May)) +
  xlab("") + 
  ylab("May Precipitation in Florida (in)")
```

Suppose you want to show box plots for each month. In this case you make the x argument in the `aes()` function the name of a column containing the vector of month names.

You first turn the data frame from its native 'wide' format to a {ggplot2} friendly 'long' format.

Wide format data is called 'wide' because it typically has a lot of columns that stretch across the screen. Long format data is called 'long' because it has fewer columns while preserving all the information. In order to do have fewer columns, it has to be longer.

Wide format data are most common. They are convenient for data entry. They let you see more of the data at one time. For example, the `FLp.df` data frame.

```{r}
head(FLp.df)
```

The long data format is less familiar. It corresponds to the relational model for storing data used by most modern databases.

Use the `pivot_longer()` function from the {tidyr} package to turn the wide data frame into a long data frame. Let's do it and then decipher what happens.

```{r}
FLpL.df <- FLp.df |>
  tidyr::pivot_longer(cols = -Year, 
                      names_to = "Month",
                      values_to = "Precipitation")

str(FLpL.df)
```

Note that the column `Month` is a character vector. When making a plot using this variable the order will be alphabetical. So instead you change it to a factor vector with levels equal to the month abbreviations.

```{r}
month.abb

FLpL.df <- FLpL.df |>
  dplyr::mutate(Month = factor(Month, 
                               levels = month.abb))

str(FLpL.df)
```

The `pivot_longer()` function takes all the columns to pivot into a longer format. Here chose them all EXCEPT the one named after the `-` sign (`Year`). All variables are measured (precipitation in units of inches) except `Year`.

The resulting long data frame has the `Year` variable in the first column and the remaining column names as the `name` variable in the second column.

You change the default `name` to `Month` by specifying the `names_to = "Month"` argument. The third column contains the corresponding precipitation values all in a single column names `value`.

You change the default `value` by specifying the `values_to = "Precipitation"`.

Note that you reverse this with the `pivot_wider()` function.

```{r}
FLpW.df <- FLpL.df |>
  tidyr::pivot_wider(id_cols = Year,
                     names_from = Month,
                     values_from = Precipitation)
```

This gif might help [gif](http://myweb.fsu.edu/jelsner/WideLong.gif).

Then to create the box plot map the `Month` variable to the x aesthetic and `Precipitation` to y aesthetic.

```{r}
ggplot(data = FLpL.df) + 
  geom_boxplot(mapping = aes(x = Month, y = Precipitation)) +
  ylab("Precipitation (in)")
```

This is a graph widely used in climatology.

Each `geom_` function is a layer. Data for the layer is specified in the function `ggplot()` with the data argument and with the mapping argument through the `aes()` function. To add another layer to the plot with different data you specify the data within the `geom_` function. For example, lets repeat the climatology graph of monthly precipitation highlighting the month of May.

You add a `geom_boxplot()` layer and specify a subset of the data using the subset `[]` operator when specifying the `data =` argument.

```{r}
ggplot(data = FLpL.df, 
       mapping = aes(x = Month, y = Precipitation)) + 
  geom_boxplot() +
  ylab("Precipitation (in)") +
  geom_boxplot(data = FLpL.df[FLpL.df$Month == "May", ], 
               mapping = aes(x = Month, y = Precipitation), 
                             fill = "green")
```

Cheat sheets: <https://www.rstudio.com/resources/cheatsheets/>

Additional examples: <https://moderndive.com/2-viz.html>

Your turn: Create a density plot of the March precipitation values in the `FLp.df` data frame. Color the density line green and label the axes.
