# Tuesday, September 12, 2022 {-}

## Today {-}

- Examples of data munging with functions from the {dplyr} package

You work with data frames. The functions are verbs. The verbs include:

Verb          | Description
-------------:|:-----------
`dplyr::select()`    | selects columns; pick variables by their names
`dplyr::filter()`    | filters rows; pick observations by their values
`dplyr::arrange()`   | reorders rows
`dplyr::mutate()`    | creates new columns; create new variables with functions of existing variables
`dplyr::summarize()` | summarizes values; collapse many values down to a single summary
`dplyr::group_by()`  | allows operations to be grouped

Syntax for the verb functions are the same:

Properties
* The first argument is a data frame. This argument is implied when using the `|>` (pipe) operator (also `%>%`).
* The subsequent arguments describe what to do with the data frame. You refer to columns in the data frame directly (without using `$`).
* The result is a new data frame

The properties make it easy to chain together simple lines of code to do something complex.

The five functions form the basis of a grammar for data. At the most basic level, you can alter a data frame in five useful ways: you can reorder the rows (`arrange()`), pick observations and variables of interest (`filter()` and `select()`), add new variables that are functions of existing variables (`mutate()`), or collapse many values to a summary (`summarise()`).

As a review consider again the Florida precipitation data set (http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt). Import the data as a data frame, _select_ the columns `April` and `Year`, _group by_ years > 1960, then _summarize_ by computing the mean and variance of the April rainfall.
```{r}
FLp.df <- readr::read_table(file = "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt")

FLp.df |>
  dplyr::select(Apr, Year) |>
  dplyr::group_by(Year > 1960) |>
  dplyr::summarize(Avg = mean(Apr),
                   Var = var(Apr))
```

## Example 1: New York City flight data {-}

Let's consider the `flights` data frame from the package {nycflights13}.
```{r}
library(nycflights13)
dim(flights)
```

The data contains all 336,776 flights that departed NYC in 2013 and comes from the U.S. Bureau of Transportation Statistics. More information is available by typing `?nycflights13`.

The object `flights` is a tibble (tabled data frame). When we have a large data frame it is useful to make it a tibble.
```{r}
head(flights)
```

The function `filter()` selects a set of rows in a data frame. How would you select all flights occurring on February 1st?
```{r}
flights |>
  dplyr::filter(month == 2 & 
                day == 1)
```

The function `arrange()` reorders the rows. If you provide more than one column name as arguments, each additional column is used to break ties in the values of the preceding columns.

How would you arrange all flights in descending order of departure delay?
```{r}
flights |>
  dplyr::arrange(desc(dep_delay))
```

Often you work with large data sets with many columns but only a few are of interest. The function `select()` allows us to zoom in on an interesting subset of the columns.

How would you create a data frame containing only the dates, carrier, and flight numbers?
```{r}
df <- flights |>
  dplyr::select(year:day, carrier, flight)
df
```

Note here the sequence operator `:` to get all columns between the column labeled `year` and the column labeled `day`.

How many distinct carriers are there? 
```{r}
df |>
  dplyr::distinct(carrier) |>
  nrow()
```

You include new columns with the function `mutate()`. Compute the time gained during flight by subtracting the departure delay (minutes) from the arrival delay.
```{r}
flights |>
  dplyr::mutate(gain = arr_delay - dep_delay) |>
  dplyr::select(year:day, carrier, flight, gain) |>
  dplyr::arrange(desc(gain))
```

Determine the average departure delay.
```{r}
flights |>
  dplyr::summarize(avgDelay = mean(dep_delay, na.rm = TRUE))
```

Note that if there are missing values in a vector the function `mean()` needs to include the argument `na.rm = TRUE` otherwise the output will be `NA`.

```{r}
y <- c(5, 6, 7, NA)
mean(y)
mean(y, na.rm = TRUE)
```

You use `sample_n()` and `sample_frac()` to take random sample of rows from the data frame. Take a random sample of five rows from the flights data frame.
```{r}
flights |>
  dplyr::sample_n(5)
```

Take a random sample of 1% of the rows.
```{r}
flights |>
  dplyr::sample_frac(.01)
```

Use the argument `replace = TRUE` to perform a _bootstrap sample_. More on this later. 

Random samples are important to modern data science.

The verbs are powerful when you apply them to groups of observations within a data frame. This is done with the function `group_by()`. Determine the average arrival delay by airplane (tail number).
```{r}
flights |>
  dplyr::group_by(tailnum) |>
  dplyr::summarize(delayAvg = mean(arr_delay, na.rm = TRUE)) |>
  dplyr::arrange(desc(delayAvg))
```

Determine the number of distinct planes and flights by destination location.
```{r}
flights |>
  dplyr::group_by(dest) |>
  dplyr::summarize(planes = dplyr::n_distinct(tailnum),
            flights = dplyr::n())
```

Repeat but arrange from most to fewest planes.

## Example 2: Daily weather data from Tallahassee {-}

Let's consider another set of data. Daily high and low temperatures and precipitation in Tallahassee. 

The file (`TLH_SOD1892.csv`) is available in this project in the folder `data`).

Import the data as a data frame.
```{r}
TLH.df <- readr::read_csv(file = "data/TLH_SOD1892.csv")
```

The data frame contains daily high (`TMAX`) and low (`TMIN`) temperatures and total precipitation (`PRCP`) from two stations: Airport with `STATION` identification USW00093805 and downtown with `STATION` identification USC00088754.

Use the `select()` function to create a new data frame with only `STATION`, `DATE`, `PRCP`, `TMAX` and `TMIN`.
```{r}
TLH.df <- TLH.df |>
  dplyr::select(STATION, DATE, PRCP, TMAX, TMIN)
TLH.df
```

Note that you've recycled the name of the data frame. You started with `TLH.df` containing all the columns and we ended with `TLH.df` with only the columns selected.

Then use the `filter()` function to keep only days at or above 90F. Similarly you recycle the name of the data frame. Use the `glimpse()` function to take a look at the resulting data frame.
```{r}
TLH.df <- TLH.df |>
  dplyr::filter(TMAX >= 90) |>
  dplyr::glimpse()
```

Note that the `DATE` column is a vector of dates having class `date`. Note if this were a character string you convert the character string into a date with the `as.Date()` function.

Functions from the {lubridate} package are used to extract information from dates. Here you add columns labeled `Year`, `Month`, and `Day` using the extractor functions `year()`, `month()`, etc.
```{r}
library(lubridate)

TLH.df <- TLH.df |>
  dplyr::mutate(Year = year(DATE),
                Month = month(DATE),
                Day = day(DATE),
                DoW = weekdays(DATE))
TLH.df
```

Next you keep only the temperature record from the airport. You use the `filter()` function on the column labeled `STATION`.
```{r}
TLH.df <- TLH.df |>
  dplyr::filter(STATION == "USW00093805")
```

Now what if you want to know how many hot days (90F or higher) by year? You use the `group_by()` function and count using the `n()` function.
```{r}
TLH90.df <- TLH.df |>
  dplyr::group_by(Year) |>
  dplyr::summarize(nHotDays = dplyr::n())

TLH90.df
```

Note that the `group_by()` function results in a data frame with the first column the variable used inside the function. In this case it is `Year`. The next columns are defined by what is in the `summarize()` function.

Repeat but this time group by `Month`.
```{r}
TLH.df |>
  dplyr::group_by(Month) |>
  dplyr::summarize(nHotDays = dplyr::n())
```

As expected the number of 90F+ days is highest in July and August. Note that you've had 90F+ days in October. 

Would you expect there to be more hot days on the weekend? How would you check this?
```{r}
TLH.df |>
  dplyr::group_by(Year, DoW) |>
  dplyr::summarize(nHotDays = dplyr::n())
```
  
You can group by more than one variable. For example, add the variable `Year` to the `group_by()` function above.

Recall that you can also `arrange()` the data frame ordered according to the values in a particular column.
```{r}
TLH90.df |>
  dplyr::arrange(desc(nHotDays))
```

## Putting everything together {-}

Let's put together your first piece of original research. You know how to import a data file, you know how to manipulate the data frame to compute something of interest, and you know how to make a graph.

Let's do this for the number of hot days. Let's say you want a plot of the annual number of hot days in Tallahassee since 1950. Let's define a hot day as one where the high temperature is at least 90F.
```{r}
library(ggplot2)

readr::read_csv(file = "data/TLH_SOD1892.csv") |>
  dplyr::filter(STATION == "USW00093805",
                TMAX >= 90) |>
  dplyr::mutate(Year = year(DATE)) |>
  dplyr::filter(Year >= 1950) |>
  dplyr::group_by(Year) |>
  dplyr::summarize(nHotDays = dplyr::n()) |>
ggplot(aes(x = Year, y = nHotDays)) +
  geom_point() +
  geom_smooth() +
  scale_y_continuous(limits = c(0, NA)) +
  ylab("Number of Days") +
  ggtitle("Number of Hot Days in Tallahassee Since 1950",
          subtitle = "High Temperature >= 90F") +
  theme_minimal()
```

You go from data in a file to a plot of interest with a set of functions that are logically ordered and easy to read.

What would you change to make a similar plot for the number of hot nights (say where the minimum temperature fails to drop below 74)?
```{r}
readr::read_csv(file = "data/TLH_SOD1892.csv") |>
  dplyr::filter(STATION == "USW00093805",
                TMIN >= 74) |>
  dplyr::mutate(Year = year(DATE)) |>
  dplyr::filter(Year >= 1950) |>
  dplyr::group_by(Year) |>
  dplyr::summarize(nHotNights = dplyr::n()) |>
ggplot(aes(x = Year, y = nHotNights)) +
  geom_point() +
  geom_smooth() +
  scale_y_continuous(limits = c(0, NA)) +
  ylab("Number of Nights") +
  ggtitle("Number of Hot Nights in Tallahassee Since 1950",
          subtitle = "Low Temperature >= 74F") +
  theme_minimal()
```

Make a similar plot showing the total precipitation by year.
```{r}
readr::read_csv(file = "data/TLH_SOD1892.csv") |>
  dplyr::filter(STATION == "USW00093805") |>
  dplyr::mutate(Year = year(DATE)) |>
  dplyr::filter(Year >= 1950) |>
  dplyr::group_by(Year) |>
  dplyr::summarize(TotalPrecip = sum(PRCP)) |>
ggplot(aes(x = Year, y = TotalPrecip)) +
  geom_point() +
  geom_smooth() +
  scale_y_continuous(limits = c(0, NA)) +
  ylab("Total Precipitation by Year") +
  theme_minimal()
```

## Example 3: Food consumption and CO2 emissions {-}

Source: https://www.nu3.de/blogs/nutrition/food-carbon-footprint-index-2018
```{r}
fc.df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
head(fc.df)
```

Consumption is kg/person/year and CO2 emission is kg CO2/person/year.

(1) How many different countries are in the data frame?
```{r}
fc.df |>
  dplyr::distinct(country) |>
  nrow()
```

(2) Arrange the countries from most pork consumption per person to the least pork consumption.
```{r}
fc.df |>
  dplyr::filter(food_category == "Pork") |>
  dplyr::select(country, consumption) |>
  dplyr::arrange(desc(consumption))
```

(3) Arrange the countries from the largest carbon footprint with respect to eating habits to the smallest carbon footprint.
```{r}
fc.df |>
  dplyr::rename(co2_emission = co2_emmission) |>
  dplyr::group_by(country) |>
  dplyr::summarize(totalEmission = sum(co2_emission)) |>
  dplyr::arrange(desc(totalEmission))
```

## Summary {-}

Data munging is a big part of data science. Data science is an iterative cycle:

1. Generate questions about our data.
2. Search for answers by _transforming_, visualizing, and modeling the data.
3. Use what you learn to refine our questions and/or ask new ones.

You use questions as tools to guide our investigation. When you ask a question, the question focuses our attention on a specific part of our data set and helps us decide what to do.

For additional practice please check out http://r4ds.had.co.nz/index.html.

Cheat sheets http://rstudio.com/cheatsheets