# Neural network example {-}

```{r}
library(neuralnet)
library(datasets)

names(infert)
```

Train the network
```{r}
nn <- neuralnet(case ~ age + parity + induced + spontaneous,
                data = infert, 
                hidden = 2,
                err.fct = "ce",
                linear.output = FALSE)
```
 
Basic results
```{r}
names(nn)

nn$result.matrix
```
 
The data is saved in `nn$covariate` and `nn$response` as well as in `nn$data` for the whole data set inclusive of non-used variables. 

The output of the neural network, i.e. the fitted values o(x), are saved
in `nn$net.result`.

```{r}
covariates.df <- as.data.frame(nn$covariate)
result.df <- data.frame(fitted = nn$net.result[[1]])
  
out.df <- cbind(covariates.df,
                result.df)
head(out.df)
```

Generalized weights
 
The generalized weight expresses the effect of each covariate has on the response and thus has an analogous interpretation as the regression parameter in regression models.

However, the generalized weight depends on all other covariates. Its distribution indicates whether the effect of the covariate is linear since a small variance suggests a linear effect
 
The columns refer to the four covariates age (j = 1), parity (j = 2), induced (j = 3), and spontaneous (j = 4)
```{r}
head(nn$generalized.weights[[1]])
```
Visualization
```{r}
plot(nn)
```