# Self-organized maps {-}

https://www.r-bloggers.com/2021/04/self-organizing-maps-in-r-supervised-vs-unsupervised/

Self-organizing maps are very useful for clustering and data visualization.

Self-organizing maps (SOMs) are a form of neural network and a way to partition complex data.

In this tutorial, we are using college admission data for clustering and visualization and we are covering unsupervised and supervised maps also.

The main objective of the tutorial is to convert high-dimensional datasets into low-dimensional maps. In others words from many variables into the two-dimensional map.

# Unsupervised Self Organizing Maps {-}

```{r}
library(kohonen)
```

```{r}
df <- readr::read_csv("data/binary.csv")
```
The data contains 400 observations and 4 variables.

You need to first normalize the data because the variables are vastly different scales.

Here you normalize with the `scale()` function where you subtract the average value from each observation and devide by the standard deviation.
