# Bivariate choropleth maps

https://cran.r-project.org/web/packages/biscale/vignettes/biscale.html

Thematic choropleth maps are used to display quantities of some variable within areas, such as mapping median income across a city’s neighborhoods. However, we often think in bivariate terms - “how do race and income vary together?”. Maps that captures this, known as bivariate choropleth maps, are often perceived as difficult to create and interpret. 

The {biscale} package has functions to implement a consistent approach to bivariate mapping.

```{r}
library(biscale)
```

All functions in the {biscale} package use the prefix `bi_` to make use of auto completion.

Consider the simple feature data frame `stl_race_income` available in the package.
```{r}
str(stl_race_income)
```
It contains the percentage of whites (`pctWhite`) and median income (`medInc`) at the census tract level for for the City of St. Louis in Missouri.

First, create a new data frame containing classes for the two variables with the `bi_class()` function. The style argument determines the type of breaks. 

The default quantile approach creates equal buckets of data for mapping, with a break created at the median (50th percentile) for a two-by-two map (`dim = 2`) or at the 33rd and 66th percentiles for a three-by-three (`dim = 3`) map.
```{r}
( df <- bi_class(stl_race_income, 
               x = pctWhite, 
               y = medInc, 
               style = "quantile", 
               dim = 3)
)
```
You can see the percentage of whites and the median income for the first tract are both in the upper third (`bi_class` is `3-3`).

Once breaks are created, you can use `bi_scale_fill()` function in a call to `ggplot()` to create a map.
```{r}
library(ggplot2)

ggplot() +
  geom_sf(data = df, 
          mapping = aes(fill = bi_class), 
          color = "white", 
          size = .1) +
  bi_scale_fill(pal = "DkBlue", dim = 3) +
  labs(title = "Race and Income in St. Louis, MO")+
  bi_theme()
```

This requires that the variable `bi_class`, created with bi_class(), is used as the fill variable in the aesthetic mapping. 

Unfortunately the does not communicate the complexity of the bi-variate scale.

To fix this you create your own legend and then add it to the map. Start by creating a ggplot map object and removing the legend.
```{r}
map <- ggplot() +
  geom_sf(data = df, 
          mapping = aes(fill = bi_class), 
          color = "white", 
          size = .1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue", dim = 3) +
  labs(title = "Race and Income in St. Louis, MO")+
  bi_theme()
```

Then you create the legend with the `bi_legend()` function.
```{r}
legend <- bi_legend(pal = "DkBlue",
                    dim = 3,
                    xlab = "Higher % White ",
                    ylab = "Higher Income ",
                    size = 5)
```

Finally you use `cowplot::ggdraw()` with `cowplot::draw_plot()` to draw the map and position the legend.
```{r}
cowplot::ggdraw() +
  cowplot::draw_plot(map, 0, 0, 1, 1) +
  cowplot::draw_plot(legend, .2, .65, .2, .2)
```