# Tornadoes over chase country

1. We hypothesize that smoother terrain is more conducive for tornado genesis especially when the nocturnal low-level jet is most pronounced. Time of day interacts with terrain roughness. Smooth terrain helps initiate tornadoes during the early evening when the low-level jet arrives.

2. The soil moisture gradient (dry-west and wet-east) portends seasonal tornado activity. Compute KBDI at two stations (Garden City-west and Salina-east) and then take the difference. Correlate the difference with tornado activity over chase country at the monthly/seasonal timescale.

Download tiles from http://viewfinderpanoramas.org/dem3.html#nam

```{r, eval=FALSE}
dem_url <- "http://viewfinderpanoramas.org/dem3/J14.zip"
download.file(dem_url,
              destfile = here::here("data", "J14.zip"))
unzip(zipfile = here::here("data", "J14.zip"),
      exdir = here::here("data"))
```

```{r}
library(terra)
Elev1.r <- terra::rast(here::here("data", "J14", "N36W099.hgt"))
# plot(Elev1.r) # check
# dim(Elev1)
Elev2.r <- terra::rast(here::here("data", "J14", "N37W099.hgt"))
Elev3.r <- terra::rast(here::here("data", "J14", "N38W099.hgt"))
Elev4.r <- terra::rast(here::here("data", "J14", "N36W100.hgt"))
Elev5.r <- terra::rast(here::here("data", "J14", "N37W100.hgt"))
Elev6.r <- terra::rast(here::here("data", "J14", "N38W100.hgt"))

Elev.m <- terra::mosaic(Elev1.r, Elev2.r, Elev3.r, 
                        Elev4.r, Elev5.r, Elev6.r)
```

```{r}
library(tmap)
tmap_mode("plot")
tm_shape(Elev.m) +
  tm_raster(palette = terrain.colors(24), alpha = .5) +
  tm_graticules(lines = FALSE)
```

Terrain roughness
```{r}
TR.m <- terra::terrain(Elev.m, 
                       v = 'roughness')

tm_shape(TR.m) +
  tm_raster(palette = "Blues", alpha = 1) +
  tm_graticules(lines = FALSE)
```

Population grid https://sedac.ciesin.columbia.edu/data/set/gpw-v4-population-density-rev11/
```{r, eval=FALSE}
file <- "gpw-v4-population-density-rev11_2020_30_sec_tif.zip"
unzip(zipfile = here::here("data", file),
      exdir = here::here("data"))
```
```{r}
Pop.r <- terra::rast(here::here("data", 
                              "gpw_v4_population_density_rev11_2020_30_sec.tif"))
Pop.r <- terra::crop(Pop.r, Elev.m)
plot(Pop.r)

tm_shape(Pop.r) + tm_raster(alpha = .3)
```

Tornado start locations
```{r}
Torn.sf <- sf::st_read(dsn = here::here("data", "1950-2020-torn-initpoint"), 
                       layer = "1950-2020-torn-initpoint")

f <- here::here("data", "1950-2020-torn-initpoint")
Torn.v <- terra::vect(f)
class(Torn.v)
dim(Torn.v)

tm_shape(log2(TR.m)) +
  tm_raster(palette = "Blues", alpha = 1) +
  tm_graticules(lines = FALSE) +
tm_shape(Torn.sf[Torn.sf$yr >= 1950, ]) +
  tm_dots() +
  tm_layout(legend.outside = TRUE)

```

Crop the tornadoes to the elevation mosaic then extract the elevation and population and terrain roughness at the tornado locations. Missing values will occur for roughness and population for locations on the borders of the domain.
```{r}
Torn.v <- terra::crop(Torn.v, Elev.m)

Elev.df <- terra::extract(Elev.m, Torn.v)
mean(Elev.df[, 2], na.rm = TRUE)

TR.df <- terra::extract(TR.m, Torn.v)
mean(TR.df[, 2], na.rm = TRUE)

Pop.df <- terra::extract(Pop.r, Torn.v)
mean(Pop.df[, 2], na.rm = TRUE)
```

Random sample of tornado locations over the domain. Background locations. https://rspatial.org/terra/sdm/index.html
```{r}
Domain.e <- terra::ext(Torn.v)
class(Domain.e)
plot(Domain.e)
```

Background data characterize the range of environmental conditions in the region. In this sense, background is the same, irrespective of where the tornadoes have originated. Background data establishes the environmental domain of the study, whilst the tornado locations establish under which conditions a tornado is more likely to be present than on average. 

```{r}
Avg = NULL
for(i in 1:1000){
Background.v <- terra::spatSample(Domain.e, 
                                  size = nrow(Torn.v),
                                  method = "random",
                                  lonlat = TRUE,
                                  as.points = TRUE)
Avg[i] <- mean(terra::extract(TR.m, Background.v)[, 2], na.rm = TRUE)
}
hist(Avg)
```

Only particular years or months
```{r}
f <- here::here("data", "1950-2020-torn-initpoint")
Torn.v <- terra::vect(f) |>
  terra::crop(Elev.m)

Torn.v <- Torn.v[Torn.v$yr >= 2001, ]
Torn.v <- Torn.v[Torn.v$mo >= 10, ]

mean(terra::extract(TR.m, Torn.v)[, 2], na.rm = TRUE)
```

Terrain roughness has a larger effect on tornado genesis during months when the large-scale dynamics is weaker (May and June).

There is not a large difference in using only the last 10-20 years.

Only particular times of the day
```{r}
f <- here::here("data", "1950-2020-torn-initpoint")
Torn.v <- terra::vect(f) |>
  terra::crop(Elev.m)

Torn.v$DateTime <- as.POSIXct(paste(Torn.v$yr, Torn.v$mo, Torn.v$dy, Torn.v$time),
                              format = "%Y%m%d%H:%M:%S")
Torn.v$Hour <- lubridate::hour(Torn.v$DateTime)

Torn.v <- Torn.v[Torn.v$Hour %in% 16:21, ]
mean(terra::extract(TR.m, Torn.v)[, 2], na.rm = TRUE)
```

Terrain roughness has a larger effect on tornado genesis during hours corresponding to when the nocturnal low-level jet occurs.
https://www.globalweatherclimatecenter.com/weather-education/what-is-the-nocturnal-low-level-jet


Create start and end spatial vectors
```{r}
f <- here::here("data", "1950-2020-torn-initpoint")
Torn.v <- terra::vect(f)

TornStart.v <- terra::vect(as.data.frame(Torn.v), 
                           geom = c("slon", "slat"),
                           crs = crs(Torn.v)) |>
  terra::crop(Elev.m)

TornEnd.v <- terra::vect(as.data.frame(Torn.v), 
                         geom = c("elon", "elat"),
                         crs = crs(Torn.v)) |>
  terra::crop(Elev.m)

mean(terra::extract(TR.m, TornStart.v)[, 2], na.rm = TRUE)
mean(terra::extract(TR.m, TornEnd.v)[, 2], na.rm = TRUE)

TornStart.v <- TornStart.v[TornStart.v$yr >= 2001, ]
TornEnd.v <- TornEnd.v[TornEnd.v$yr >= 2001, ]

mean(terra::extract(TR.m, TornStart.v)[, 2], na.rm = TRUE)
mean(terra::extract(TR.m, TornEnd.v)[, 2], na.rm = TRUE)
```

Create a single data frame with tornado locations (labeled 1) and background locations (labeled 0) that includes elevation, population, and terrain roughness.
```{r}
f <- here::here("data", "1950-2020-torn-initpoint")
Torn.v <- terra::vect(f) |>
  terra::crop(Elev.m)

Torn.v <- Torn.v[Torn.v$mag >= 1, ]

Elev.df <- terra::extract(Elev.m, Torn.v)
TR.df <- terra::extract(TR.m, Torn.v)
Pop.df <- terra::extract(Pop.r, Torn.v)
Torn.df <- data.frame(Location = 1, 
                      Elevation = Elev.df[, 2], 
                      Roughness = TR.df[, 2], 
                      Population = Pop.df[, 2])

Background.v <- terra::spatSample(Domain.e, 
                                  size = nrow(Torn.v),
                                  method = "random",
                                  lonlat = TRUE,
                                  as.points = TRUE)
Elev.df <- terra::extract(Elev.m, Background.v)
TR.df <- terra::extract(TR.m, Background.v)
Pop.df <- terra::extract(Pop.r, Background.v)
Background.df <- data.frame(Location = 0, 
                            Elevation = Elev.df[, 2], 
                            Roughness = TR.df[, 2], 
                            Population = Pop.df[, 2])

TornBack.df <- rbind(Torn.df, Background.df)
```

Logistic regression
```{r}
model1 <- glm(Location ~ Population + Elevation + Roughness,
              family = "binomial", data = TornBack.df)
summary(model1)

predGrid <- data.frame(Population = 1500, 
                       Elevation = 790,
                       Roughness = 20)

predict(model1, newdata = predGrid, type = "response")
```

```{r}
remotes::install_github("rspatial/predicts")

model2 <- glm(Location ~ Elevation + Roughness,
              family = "binomial", data = TornBack.df)
summary(model2)


predictors <- c(Elev.m, TR.m)
names(predictors) <- c("Elevation", "Roughness")

pg <- predict(predictors, model2, type = "response")
plot(pg)
```

Create supercells https://github.com/Nowosad/supercells
```{r}
pg1 <- supercells::supercells(pg, k = 100, compactness = .2)

plot(pg)
plot(st_geometry(pg1), add = TRUE)

tm_shape(pg1) +
  tm_fill(col = "lyr1")
```

Daily weather data
```{r}
Salina.df <- readr::read_csv(file = here::here("data", "Salina_Kansas.csv")) |>
  dplyr::mutate(Year = lubridate::year(DATE)) |>
  dplyr::filter(Year >= 2000) |>
  dplyr::select(DATE, Year, PRCP, TMAX, TMIN, TAVG)

GardenCity.df <- readr::read_csv(file = here::here("data", "GardenCity_Kansas.csv"))


```
https://www.ncei.noaa.gov/maps/daily-summaries/