# Self-organized maps {-}

https://www.r-bloggers.com/2021/04/self-organizing-maps-in-r-supervised-vs-unsupervised/

Self-organizing maps are very useful for clustering and data visualization.

Self-organizing maps (SOMs) are a form of neural network and a way to partition complex data.

In this tutorial, we are using college admission data for clustering and visualization and we are covering unsupervised and supervised maps also.

The main objective of the tutorial is to convert high-dimensional datasets into low-dimensional maps. In others words from many variables into the two-dimensional map.

## Unsupervised Self Organizing Maps {-}

```{r}
library(kohonen)
```

```{r}
df <- readr::read_csv("data/binary.csv")
```
The data contains 400 observations and 4 variables.

You need to first normalize the data because the variables are vastly different scales.

Here you normalize with the `scale()` function where you subtract the average value from each observation and divide by the standard deviation.

```{r}
dfS <- df |>
  dplyr::mutate(greS = scale(gre),
                gpaS = scale(gpa),
                rankS = scale(rank)) |>
  dplyr::select(greS, gpaS, rankS) |>
  as.matrix()
summary(dfS)
```
The scaled variables all have a mean of zero.

First you create a 4 x $ grid with the `somgrid()` function.
```{r}
set.seed(222)
g <- somgrid(xdim = 4, 
             ydim = 4, 
             topo = "rectangular" )
```

Next you map the variables to the grid with the `som()` function.
```{r}
map <- som(X = dfS,
           grid = g,
           alpha = c(.05, .01), 
           radius = 1)
```

`alpha` is the learning weight by default vale is 0.05 to 0.01. These two numbers basically indicate amount of change.

Plot
```{r}
plot(map, 
     type = 'codes',
     palette.name = rainbow)
```

These provides codes plot with rainbow colors.

For example, first node indicates higher gre values compared to other variables.

```{r}
map$unit.classif
```

```{r}
map$codes
```

```{r}
plot(map, type = "mapping")
```

## Supervised SOMS

We need to split the dataset into train and test data sets for the prediction and accuracy checking.

Create independent samples first.
```{r}
set.seed(123)
ind <- sample(2, nrow(df), replace = T, prob = c(0.7, 0.3))
m <- as.matrix(df)
train <- m[ind == 1,]
test <- m[ind == 2,]
```

The training dataset contains 285 observations and the test has 115  observations.

You again need to normalize the variables.
```{r}
trainX <- scale(train[,-1])
testX <- scale(test[,-1],
               center = attr(trainX, "scaled:center"),
               scale = attr(trainX, "scaled:scale"))
trainY <- factor(train[,1])
Y <- factor(test[,1])
test[,1] <- 0
testXY <- list(independent = testX, dependent = test[,1])
```

Gradient Boosting in R
```{r}
set.seed(223)
map1 <- xyf(trainX,
            classvec2classmat(factor(trainY)),
            grid = somgrid(5, 5, "hexagonal"),
            rlen = 100)
plot(map1, type='codes', palette.name = rainbow)
```

You create cluster boundaries and plot both the graphs.

```{r}
par(mfrow = c(1,2))
plot(map1,
     type = 'codes',
     main = c("Codes X", "Codes Y"))
map1.hc <- cutree(hclust(dist(map1$codes[[2]])), 2)
add.cluster.boundaries(map1, map1.hc)
par(mfrow = c(1,1))
```

Prediction
```{r}
pred <- predict(map1, newdata = testXY)
```

Now let’s see the misclassification error based on above model.
```{r}
table(Predicted = pred$predictions[[2]], Actual = Y)
```
Based on the confusion matrix, total of 58 + 12 = 70 correct classifications and 45 misclassifications. So, it indicates that the model accuracy is that we get here is 61% (70/(70+45)).