---
title: "Assignment 1"
author: "Answers"
output: html_document
---

Due Date: September 1, 2022 (8 p.m.)

Total Points: 35

**1** (5) Change the name of the Rmd file to `your_last_name_A1.Rmd`.

**2** (5) Inside the Rmd file, change `Your Full Name` above to your name.

**3** (5) Write a sentence describing what you want to do after you graduate.

I would like to become a professor.

**4** (5) Insert the code delimiters (back ticks and curly braces) and then, between the delimiters, type `library(ggplot2)` on one line and `mpg` on the next line.

```{r}
library(ggplot2)
mpg
```


**5** (5) Insert code that (a) lists the column names of the data frame `mpg` and (b) computes the average city mileage (in miles per gallon).

```{r}
names(mpg)
mean(mpg$cty)
```

**6** (5) Using the data frame `mpg` create a scatter plot showing city mileage on the horizontal axis and highway mileage on the vertical axis. 

```{r}
ggplot(data = mpg,
       mapping = aes(x = cty, y = hwy)) +
  geom_point()
```


**7** (5) Knit this Rmd file to create an HTML file.

Email your finished assignment file (the Rmd file) to [jelsner\@fsu.edu](mailto:jelsner@fsu.edu){.email}
